Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/figures/splitting-overview.tex

\put(28,66){\parbox[t]{44 \unitlength}{ 
	\textbf{Correctness Condition:} \\ 
	Splitting followed by evaluation (top and right arrows)
	produce the same result as 
	evaluation followed by masking (left and bottom arrows),
	up to a term equivalence.
}}
\end{picture}

@@

\put(28,66){\parbox[t]{44 \unitlength}{ 
	\textbf{Correctness Condition:} \\ 
	Splitting followed by evaluation (top and right arrows)
	produce the same result as 
	direct evaluation (left and bottom arrow),
	up to a term equivalence.
}}
\end{picture}

>>>>

\put(28,66){\parbox[t]{44 \unitlength}{ 
	\textbf{Correctness Condition:} \\ 
	Splitting followed by evaluation (top and right arrows)
	produces the same result as 
	\lang\ evaluation followed by masking (left and bottom arrows),
	up to a term equivalence.
}}
\end{picture}

@@

\put(28,66){\parbox[t]{44 \unitlength}{ 
	\textbf{Correctness Condition:} \\ 
	Splitting followed by evaluation (top and right arrows)
	produces the same result as 
	\lang\ evaluation (left and bottom arrow),
	up to a term equivalence.
}}
\end{picture}

### 2015-02-27 paper/semantics.tex

Although products and functions are restricted to types at the same world,
$\fut$ allows costruction of ``mixed-stage'' products and functions.
For example, quickselect is a function at world \bbonem\ which takes a
$\curr{\rm list} \times \fut\rmint$ (a purely-first-stage list and a
second-stage integer) to a $\fut\rmint$ (a second-stage
computation of an integer).

@@

>>>>

Although products and functions are restricted to types at the same world,
$\fut$ allows construction of ``mixed-stage'' products and functions.
For example, quickselect is a function at world \bbonem\ which takes a
$\curr{\rm list} \times \fut\rmint$ (a purely-first-stage list and a
second-stage integer) to a $\fut\rmint$ (a second-stage
computation of an integer).

@@

Although the type system of \lang\ (specifically, the opaque nature of $\fut$
types) ensures that we can perform all first-stage computations before any
second-stage ones, the fact that first- and second-stage computations are
interleaved in \lang\ terms means that realizing this notion of staged
evaluation requires some care.

### 2015-02-27 paper/splitting.tex

Splitting statically transforms terms in \lang\ into a pair of
monostaged terms that correspond to the first stage and second stage
computations in the original term. The idea is that evaluating the
first monostaged term in a first pass, and then evaluating the second
monostaged term on the results of the first pass, yields the same
result as fully evaluating the original \lang\ program.  Complexity
arises because \lang\ evaluation of \bbonem\ terms may also generate
results in the first stage. (For example, a \bbonem\ term of type
$\curr \rmint$ results in an \rmint\ in the first stage.) Handling the
asymmetry between the results of evaluating \bbonem\ terms (which may
produce results at both the first and second stage) and \bbtwo\ terms
(which only produce second stage results) is fundamental to the
structure of the splitting algorithm presented in this section.  By
construction, \bbonep\ terms are monostage and do not require
splitting.

@@

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms $c$
and $l.r$
%(corresponding to the first stage and second stage
%computations in $e$).
Evaluating $c$ in the first pass produces the pair $(i,b)$, where the
first element $i$ matches the first stage results contained in $v$.
For example, in the simple case where $\rtab \xi v$ contains entirely
first stage computation, such as:
\begin{lstlisting}
2`[]`1` grnd{1+2}`
\end{lstlisting}
the first stage result is the value 3.  The second element of the pair
is used as input to the resumer in second pass evaluation as $\reduce
{[b/l]r} {v'}$, where $v'$ is the same as result produced by second
stage computations in $e$.  For example, in the case where $\rtab \xi
v$ contains entirely second stage computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into stage one and stage two results. We then describe
the algorithm for splitting both \bbonem\ terms ($\splitonesym$) and
\bbtwo\ terms ($\splittwosym$).

>>>>

Splitting statically transforms terms in \lang\ into a pair of
monostaged terms that correspond to the first stage and second stage
computations in the original term. The idea is that evaluating the
first monostaged term in a first pass, and then evaluating the second
monostaged term on the results of the first pass, yields the same
result as fully evaluating the original \lang\ program (performing
both first stage and second stage evaluation).  While world
\bbtwo\ terms fully evaluate to values in stage two, complexity arises
because evaluation of \bbonem\ terms may also generate results in the
first stage. (For example, a \bbonem\ term of type $\curr \rmint$
results in an \rmint\ in the first stage.) Handling this asymmetry
between the results of evaluating \bbonem\ terms (which may produce
results at both the first and second stage) and \bbtwo\ terms (which
only produce second stage results) is fundamental to the structure of
the splitting algorithm presented in this section.  By construction,
\bbonep\ terms are monostage and do not require splitting.

@@

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms $c$
and $l.r$. Evaluating $c$ in the first pass produces the pair $(i,b)$,
where the first element $i$ matches the first stage results contained
in $v$.  For example, in the simple case where $\rtab \xi v$ contains
entirely first stage computation, such as:
\begin{lstlisting}
2`[]`1` grnd{1+2}`
\end{lstlisting}
the first stage result is the value 3.  The second element of the pair
is used as input to the resumer in second pass evaluation as $\reduce
{[b/l]r} {v'}$, where $v'$ is the same as result produced by second
stage computations in $e$.  For example, in the case where $\rtab \xi
v$ contains entirely second stage computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into stage one and stage two results. We then describe
the algorithm for splitting both \bbonem\ terms ($\splitonesym$) and
\bbtwo\ terms ($\splittwosym$).
